The illusory promise of the Aligned Rank Transform

### 2025-07-08 text/20-background.qmd
minor correction
corrected links

Even in this very small dataset ($n = 2$), the three ART rankings are distinct from one another and also differ significantly from the ranking produced by a simple rank transformation. Interestingly, identical responses (e.g., the two values of $0.3$) can be assigned very different ranks, while responses that differ substantially (e.g., $0.3$ and $3.0$) may receive the same rank. More surprisingly, larger values can receive lower ranks —-- revealing that ART is a non-monotonic transformation. For instance, for subject $s_1$ under condition $b_1$, the responses are $0.4$ and $0.8$ for $a_1$ and $a_2$, yet their respective $ART_A$ ranks are $7.5$ and $3.0$. In this case, ART significantly distorts the actual difference between the two groups, $a_1$ and $a_2$.

@@

Out of the 39 papers examined, 38 identify at least one statistically significant main effect using ART. Additionally, 30 papers conduct tests for interactions, with 24 of them reporting at least one statistically significant interaction effect. Only five papers have publicly available data. We revisit the results of three of these papers in [Section 6](#casestudies).

@@

With a focus on clarity, we chose to exclusively examine the three rank-based transformations presented earlier: the pure rank transformation (RNK), ART, and INT. While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in the [appendix](appendix.html). These findings indicate that its performance is comparable to the rank transformation but seems to be inferior to the simpler and more versatile INT. The appendix features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests. Our findings show that, at least in balanced designs, these methods suffer from low power issues in certain situations without demonstrating any clear benefits compared to INT.

Finally, we chose to limit our investigation to balanced experimental designs. The rationale behind this decision is that we have not encountered any prior claims about ART's suitability for unbalanced data, and the ARTool [@artool] issues a warning in such situations. However, the [appendix](appendix.html) presents additional results on experimental designs with missing data, which confirm that in such situations, ART's accuracy may deteriorate further.

>>>>

Even in this very small dataset ($n = 2$), the three ART rankings are distinct from one another and also differ significantly from the ranking produced by a simple rank transformation. Interestingly, identical responses (e.g., the two values of $0.3$) can be assigned very different ranks, while responses that differ substantially (e.g., $0.3$ and $3.0$) may receive the same rank. More surprisingly, larger values can receive lower ranks --- revealing that ART is a non-monotonic transformation. For instance, for subject $s_1$ under condition $b_1$, the responses are $0.4$ and $0.8$ for $a_1$ and $a_2$, yet their respective $ART_A$ ranks are $7.5$ and $3.0$. In this case, ART significantly distorts the actual difference between the two groups, $a_1$ and $a_2$.

@@

Out of the 39 papers examined, 38 identify at least one statistically significant main effect using ART. Additionally, 30 papers conduct tests for interactions, with 24 of them reporting at least one statistically significant interaction effect. Only five papers have publicly available data. We revisit the results of three of these papers in [Section 7](#casestudies).

@@

With a focus on clarity, we chose to exclusively examine the three rank-based transformations presented earlier: the pure rank transformation (RNK), ART, and INT. While we do not elaborate on the performance of ATS in the main paper, additional experimental results are available in the [appendix](appendix.html#ATS). These findings indicate that its performance is comparable to the rank transformation but seems to be inferior to the simpler and more versatile INT. The [appendix](appendix.html#generalized) features additional results regarding the performance of the generalized van der Waerden test, as well as the generalized Kruskal-Wallis and Friedman tests. Our findings show that, at least in balanced designs, these methods suffer from low power issues in certain situations without demonstrating any clear benefits compared to INT.

Finally, we chose to limit our investigation to balanced experimental designs. The rationale behind this decision is that we have not encountered any prior claims about ART's suitability for unbalanced data, and the ARTool [@artool] issues a warning in such situations. However, the [appendix](appendix.html#missing) presents additional results on experimental designs with missing data, which confirm that in such situations, ART's accuracy may deteriorate further.
